The contact value approximation for the pair distribution function of the inhomogeneous hard-sphere fluid

### 2014-03-12 papers/pair-correlation/paper.tex

\begin{document}
\title{The Pair Distribution Function of the Inhomogeneous Hard-Sphere
  Fluid}

@@

%\pacs{61.20.Ne, 61.20.Gy, 61.20.Ja}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We have developed an approximation for the pair distribution
  function of hard spheres, which is based on the average value of the
  pair distribution function at contact we recently developed.  Our
  pair distribution function is constructed so as to be computable
  with only single-site convolutions and to accurately reproduce the
  averaged distribution function at contact.  This results in a
  functional that is efficient to compute, and predicts the pair
  distribution function as well as the most accurate existing
  analytical approximation.  \fixme{needs special attention.}
\end{abstract}

@@

\section{Our approach}
%% We approximate the pair correlation with the average of two radial
%% distribution functions, evaluated at the distance between the two
%% points, that are themselves functions of the averaged pair correlation
%% at contact $g_{\sigma}(\rr)$ discussed above~\cite{schulte2012using},
%% evaluated at the two points.
In the approaches for the pair distribution function mentioned above,
the radial distribution function used to approximate pair correlation
was dependent upon the density averaged over some volume.  We seek to
achieve greater accuracy by using instead a function dependent upon
our averaged $g_{\sigma}(\rr)$ discussed above, which holds more
information about an inhomogeneous system than does a simple
convolution of the density.
%
We approximate the pair correlation with the average of two radial
distribution functions, evaluated at the distance between the two
points, that are themselves functions of the averaged pair correlation
at contact $g_{\sigma}(\rr)$ evaluated at the two points:
%
%% We therefore seek a function form
%% for our radial distribution that is dependent upon $g_{\sigma}(\rr)$
%% and that will approach the correct solution in the homogeneous limit
%% and also in the limit of contact distance between the two points.
\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) +
    g(|r_{12}|, g_\sigma(\rr_2))}{2}. \label{eq:g2-our-mean}
\end{align}
$g^{(2)}(\rr_1,\rr_2)$ above is constructed to reproduce the exact
value for the integral$ \int n(\rr_1)n(\rr_2)
g^{(2)}(\rr_1,\rr_2)\delta(|\rr_1-\rr_2|-\sigma)d\rr_1d\rr_2$.  \par
The approximation described in Eq.~\ref{eq:g2-our-mean} requires the
radial distribution function as a function of $r$ and $g_\sigma$.  One
approach to constructing $g(r,g_\sigma)$ is to compute $\eta$ from
$g_\sigma$ and then find $g(r,\eta)$ either using some approximate
functional form (e.g. Percus-Yevick) or using Monte Carlo simulation
results.  In this paper we first use the Monte Carlo results (which
are also used in evaluating the approximations of earlier works), and
refer to this approximation (which has an exact $g(r,g_\sigma)$) as
``MCRD.''  \jeffsays{I thought maybe for Monte Carlo Radial
  Distribution} We will then construct our own, more computationally
efficient radial distribution approximation, which is addressed in the
following section.

@@

\subsection{Separable fit to the radial distribution function}

The approximation described in Eq.~\ref{eq:g2-our-mean} is reasonably
accurate, but cannot be computed efficiently with single-center
convolutions as written.  In order to ease computational application
of this pair distribution function in classical density functionals,
we have developed a separable appoximation for the function
$g(r,g_\sigma)$.
\begin{align}
  g(|r_{12}|, g_\sigma) &= \sum_{i} a_i(r_{12}) b_i(g_\sigma)
\end{align}
This separable approximation allows the mean field contribution to
the free energy (Eq.~\ref{eq:mean-field}) to be computed with only
single-center convolutions, which can be efficiently computed using
fast Fourier transforms.

@@

A particularly relevant quantitative test of a pair distribution
function is how well it predicts the interaction energy due to a pair
potential.  To this end, we have computed the error in the first term
in a high-temperature perturbation expansion $a_1$ for several pair
potentials.  In order to focus on effects at the interface, we have
defined a position-dependent pair interaction energy as
\begin{align}
  \frac{dF_1}{dz} &=
  \tfrac12 \int g^{(2)}_{HS}(\rr,\rr')n(\rr)n(\rr')\Phi(|\rr-\rr'|)
  d\rr'\, dxdy\label{eq:da1}
\end{align}
which gives the contribution to the mean-field free energy due to
molecules located at position $z$.  \jeffsays{About this last
  sentence: this is molecules in the plane $z$, right?  Wanted to make
  sure that's the case before changing to something like 'in the
  plane'.  Saying the position z makes it seem that there's only one.}

>>>>

\begin{document}
\title{The contact value approximation for the pair distribution function of the inhomogeneous hard-sphere
  fluid}

@@

%\pacs{61.20.Ne, 61.20.Gy, 61.20.Ja}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We have developed an approximation for the pair distribution
  function of hard spheres, which is based on the average value of the
  pair distribution function at contact we recently developed.  Our
  pair distribution function is constructed so as to be computable
  with only single-site convolutions and to accurately reproduce the
  averaged distribution function at contact.  This results in a
  functional that is efficient to compute, and predicts the pair
  distribution function as well as the most accurate existing
  analytical approximation.  \fixme{needs special attention.}
  \fixme{add acronym, and discuss separable vs. non}
\end{abstract}

@@

\section{Contact value approximation}
%% We approximate the pair correlation with the average of two radial
%% distribution functions, evaluated at the distance between the two
%% points, that are themselves functions of the averaged pair correlation
%% at contact $g_{\sigma}(\rr)$ discussed above~\cite{schulte2012using},
%% evaluated at the two points.
In the approaches for the pair distribution function mentioned above,
the radial distribution function used to approximate pair correlation
was dependent upon the density averaged over some volume.  We seek to
achieve greater accuracy by using instead a function dependent upon
our averaged $g_{\sigma}(\rr)$ discussed above, which holds more
information about an inhomogeneous system than does a simple
convolution of the density.
%
We approximate the pair correlation with the average of two radial
distribution functions, evaluated at the distance between the two
points, that are themselves functions of the averaged pair correlation
at contact $g_{\sigma}(\rr)$ evaluated at the two points:
%
%% We therefore seek a function form
%% for our radial distribution that is dependent upon $g_{\sigma}(\rr)$
%% and that will approach the correct solution in the homogeneous limit
%% and also in the limit of contact distance between the two points.
\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) +
    g(|r_{12}|, g_\sigma(\rr_2))}{2}. \label{eq:g2-our-mean}
\end{align}
$g^{(2)}(\rr_1,\rr_2)$ above is constructed to reproduce the exact
value for the integral$ \int n(\rr_1)n(\rr_2)
g^{(2)}(\rr_1,\rr_2)\delta(|\rr_1-\rr_2|-\sigma)d\rr_1d\rr_2$.  \par
The approximation described in Eq.~\ref{eq:g2-our-mean} requires the
radial distribution function as a function of $r$ and $g_\sigma$.  One
approach to constructing $g(r,g_\sigma)$ is to compute $\eta$ from
$g_\sigma$ and then find $g(r,\eta)$ either using some approximate
functional form (e.g. Percus-Yevick) or using Monte Carlo simulation
results.  In this paper we first use the Monte Carlo results (which
are also used in evaluating the approximations of earlier works), and
refer to this approximation (which has an exact $g(r,g_\sigma)$) as
the ``contact value approximation'' (CVA).  \fixme{Should we mention
  this earlier?}

@@

\section{Separable radial distribution function for the CVA}

The contact value approximation (described in
Eq.~\ref{eq:g2-our-mean}) is reasonably accurate, but cannot be
computed efficiently with single-center convolutions as written.  In
order to ease computational application of this pair distribution
function in classical density functionals, we have developed a
separable appoximation for the function $g(r,g_\sigma)$.
\begin{align}
  g(|r_{12}|, g_\sigma) &= \sum_{i} a_i(r_{12}) b_i(g_\sigma)
\end{align}
This separable contact value approximation (CVA-S) allows the mean
field contribution to the free energy (Eq.~\ref{eq:mean-field}) to be
computed with only single-center convolutions, which can be
efficiently computed using fast Fourier transforms.  \fixme{Comb paper
  for places to say CVA-S and CVA.}

@@

A particularly relevant quantitative test of a pair distribution
function is how well it predicts the interaction energy due to a pair
potential.  To this end, we have computed the error in the first term
in a high-temperature perturbation expansion $a_1$ for several pair
potentials.  In order to focus on effects at the interface, we have
defined a position-dependent pair interaction energy as
\begin{align}
  \frac{dF_1}{dz} &=
  \tfrac12 \int g^{(2)}_{HS}(\rr,\rr')n(\rr)n(\rr')\Phi(|\rr-\rr'|)
  d\rr'\, dxdy\label{eq:da1}
\end{align}
which gives the contribution to the mean-field free energy due to
molecules located in a plane of fixed $z$.
